---
title: "Choose Your Own Project"
author: "Greg Baguhin"
date: "28/07/2021"
output: pdf_document
---
#### INTRODUCTION

This project attempts to fit a linear & logistical regression model to a clothes size dataset. It will try to correlate the effectiveness of the predictors in giving accurate predictions for cloth size.

##### OBJECTIVE: 

I will be using the cloth_size dataset downloaded from Kaggle to identify the most significant predictor that would produce the most accurate cloth size for a given variable.


#### DATA PREPARATION

Splitting the dataset into training and testing sets at a 50-50 ratio. This was chosen due to the relatively small amount of data used so the test set can be just as random as the training set.

```{r message=FALSE,warning=FALSE}
#install.packages("tidyverse")
library(tidyverse)
#install.packages("tidyr")
library(tidyr)
#install.packages("caret")
library(caret)


#Data Preparation:
# Loading the csv file into RStudio
url <- "D:/Desktop/Harvard Online/Data Science/Capstone/archive/cloth_size.csv"
clothsize_data <- read_csv(url)

# View first 6 rows
head(clothsize_data)

# inspect data properties
str(clothsize_data)

# inspect data for statistics
summary(clothsize_data)

# Find out how many rows with NAs
sum(is.na(clothsize_data))

# converting size to a factor
clothsize_data <- clothsize_data %>% mutate(size=as_factor(size))

# define the outcome and predictors
y <- clothsize_data$size
a <- clothsize_data$age
b <- clothsize_data$weight
c <- clothsize_data$height


# generate training and test sets
set.seed(1, sample.kind = "Rounding") # if using R 3.5 or earlier, remove the sample.kind argument
test_index <- createDataPartition(y, times = 1, p = 0.5, list = FALSE)
test_set <- clothsize_data[test_index, ]
train_set <- clothsize_data[-test_index, ]
```

##### LINEAR REGRESSION MODEL

Let's look at the predictor averages for each size

```{r warning=FALSE}
clothsize_data %>% group_by(size) %>%
  summarise(avg_wt = mean(weight), avg_ht = mean(height), avg_age = mean(age))
```


Plotting probability of a size "M" vs weight
```{r message=FALSE, warning=FALSE}
clothsize_data %>%
  mutate(x = round(weight)) %>%
  group_by(x) %>%
  filter(n() >= 10) %>%
  summarize(prop = mean(size == "M")) %>%
  ggplot(aes(x, prop)) +
  geom_point()
```

We see that a person who is 58kgs has a 50% probability of wearing a size "M"

```{r message=FALSE, warning=FALSE}
train_set %>%
  filter(round(weight)==58) %>%
  summarize(y_hat = mean(size=="M"))
```


##### LOGICTIS REGRESSION MODEL

Fit logistic regression model using weight as a predictor for size M, the size with
the highest frequency in the dataset.

```{r message=FALSE, warning=FALSE}
glm_fit <- train_set %>% 
  mutate(y = as.numeric(weight==58)) %>%
  glm(y ~ size, data=., family = "binomial")

p_hat_logit <- predict(glm_fit, newdata = test_set, type = "response")


```

#### CONCLUSION:

We were able to show that we can use linear & logistic regression to model predictions for cloth size using the dataset. A more thorough testing algorithm could be built in future projects applying the same principles learned from this project.

